|

### 2020-09-22 paper.Rmd
merging paragraphs with crowding problem, expands on structure

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections, like the tour [@As85,@BCAH05]. The effect is described by @diaconis1984, that most low-dimensional linear projections will show approximately Gaussian distributions of the data, with observations concentrating in the center. This motivated the development of indexes for projection pursuit searching for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations are in a single point in certain low-dimensional projections.  These issues also persist with non-linear dimension reduction techniques, and referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

Figure \ref{fig:density} illustrates the crowding problem. Two-dimensional linear projections of points sampled uniformly within $p$-dimensional hyperspheres in $p=3, 10, 100$ are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection.

@@

In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods. Tours show interpolated sequences of low-dimensional projections of the data. When exploring a dataset with a tour we can discover features that are only visible from specific viewing angles. However, these features could be hidden in the observation pile, and even worse leaving a large fraction of the available plotting canvas sparsely populated. To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution. This is called a burning sage tour, to reflect that the curse of dimensionality is being removed.

The paper is structured as follows. The radial transformation and its implementation is described in Section \ref{sec:method}. Section \ref{sec:application} illustrates the use of the sage tour with examples in clustering, supervised classification and a classic needle-in-the-haystack problem.

>>>>

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections, like the tour [@As85,@BCAH05]. The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center. This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate. Figure \ref{fig:density} illustrates the crowding problem. Two-dimensional linear projections of points sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection.

@@

In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods. Tours show interpolated sequences of low-dimensional projections of the data. When exploring data with a tour we can discover features that are only visible in linear combinations of variables. However, the data crowding could hide these features. To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution. This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.

The paper is structured as follows. The radial transformation and its implementation is described in Section \ref{sec:method}. Section \ref{sec:application} illustrates the use of the sage tour with examples in clustering, supervised classification and a classical needle-in-the-haystack problem. Section \ref{sec:concl} describes possible extensions to the method.
